Arrows for Parallel Computation

### 2019-12-27 paper/content/response.md

Thank you.

> But when approaching the final steps of the paper one question became 
> increasingly pending:  So what?
> 
> What is the real underlying contribution of the paper, in the sense of novel 
> insight, beyond the foreground contributions summarised in the paper?
>

TODO

Our formalism inspired S. Zhang's Session Arrows (Zhang, 2019).

@@

> 
> page 5:
> 
> Like on page 2, the diversion towards Frege remains cryptic.
>

We clarified the wording.

> page 7:
> 
> Provide a brief characterisation of profunctors, or at least a reference.
> "since they they allow"

We rephrased the offending paragraph, providing an additional reference explaining Profunctors (Boisseau & Gibbons, 2018). The text emphases now that it is possible to define Arrows deriving from Profunctors, as Appendix B shows. There we also cite Asada (2010).

> 
> The introductory text of section 3.2 (before 3.2.1) is mostly re-iterating 
> aspects already discussed in previous sections, partially with identical
> phrasing. I suggest to cut this short.

The text was removed.

> 
> Other side issues like LVish or HdpH should have been discussed earlier.
> The background section should really focus on the relevant background,
> whereas the selection of background topics should be clear at this point.
>

We moved the text ob LVish and HdpH to the related work.

> page 10:
> 
> "efficiency OR the lack thereof" maybe?

Fixed.

> 
> page 11:
> 
> The dashes should be longer, or in Latex terms: "---".

Fixed globally.

> 
> page 13:
> 
> "instances of the ArrowParallel instances" ?

This was obviously a writing mishap, fixed.

> 
> page 15:
> 
> What is meant be the term "implementation agnostic"? Any (high-level) 
> programming approach is somewhat implementation-agnostic.

We clarified the wording and emphasised portability across parallel Haskells.

> 
> page 16ff:
> 
> The text flow continuously jumps to the appendix and back; numerous forward
> references to figures in the appendix are increasingly annoying for the 
> reader. I believe that the authors should make a choice: if some example,
> implementation, illustration, etc, is relevant for the reader then it should
> appear immediately and not in the appendix. Everything else should either be
> skipped or the remaining contents of the appendix should be explained once.
>

> page 16:
> 
> "toplogical"

Fixed.

> 
> page 19:
> 
> The question arising on the top of page 19 should be investigated. The statement
> that such differences would be beyond the scope of the paper appears a bit lame.
>

We changed the wording.

Our reasoning is that the behaviour in question arises from the subtle details of laziness and demand in Eden standard library. We argue that such details are not of major interest to the generic JFP audience. Our paper focuses on establishing a common interface between different parallel Haskells, with a minor workaround the above demand issue becomes irrelevant.

> page 22:
> 
> "repeating": why is the front part of the word set in italics?

We were referencing the `repeat` function. We now use an apostrophe for more clarity.

> 
> page 23:
> 
> "using the Gentleman algorithm" -> "using Gentleman's algorithm"

Fixed.

> 
> page 24:
> 
> If the purpose is to compare Figures 29 and 30, I strongly suggest to make
> sure they appear on the same page, or better the graphs are combined in a 
> single figure. A related question is: why should there be any observable
> difference? After all the exact same infrastructure and almost the same
> implementation are used.

We forced both figures on a single page.

The point of those figures is exactly to showcase that there is no observable difference, despite using a further intermediate abstraction layer in form of our API.

> 
> page 25:
> 
> I find the statement that "PArrows hold up in terms of performance when 
> compared with the original parallel Haskells" a bit strong. At the end
> of the day, PArrows is not a new parallel Haskell, but merely an abstraction
> wrapper around the existing parallel Haskells. 
> 
> It is of little surprise that 32 hyperthreaded cores do not behave exactly as
> 16 real cores. Still the conclusion to disregard hyperthreading because it
> behaves differently appears illogic. I'm inclined to agree that hyperthreading
> might not provide any particular interesting insights for the questions
> concerned, but the justification should be rephrased.
> 
> page 27:
> 
> The text in 7.1.4 and the beginning of 7.2 should be condensed.

We implemented this change request.

> 
> page 28:
> 
> A graphical illustration of the experimental findings appears mandatory.
> The mentioning of a zillion number is pretty unreadable and complicates
> comprehension of the results. The text should be used to explain the most
> interesting numbers and discuss the findings.

> 
> The term "speedup" is used extensively without ever explaining the base line.

We added a definition of speedup.

> 
> page 30:
> 
> The absolute runtimes on 16 cores are fairly small. This raises the question
> whether larger problem sizes or simply repeated computing of the existing
> problem sizes could contribute to noise reduction and might possibly lead to
> more reliable experimental results.
> 
> "extensible formalism THAT can be easily ported"

Fixed.

> 
> The sentence "We argue that ..." is pretty much repeated right thereafter.

We removed this sentence all together.

> 
> page 32ff:
> 
> The references contain numerous capitalisation errors. For conferences
> there is an unmotivated but consistent space between the acronym and the year.

The space is removed.

> 
> 
> ***************************************************************************
> 
> 
> 
> 
> Referee: 2
> 
> Comments to the Author
> Summary:
> 
> This paper introduces the use of Arrows for specifying parallelism. In
> particular, it defines an Arrow-based domain specific language
> (library) and implements it using multiple parallel Haskells.
> 
> More specifically, it introduces specialized Arrows as a general
> interface to parallel computations, by defining the ArrowParallel 
> type class and giving instance implementations for GpH, the Par Monad 
> and Eden. To demonstrate its expressiveness, it shows that the
> Parallel Arrows can be used to define some useful algorithmic
> skeletons that abstract typical parallel computations. The
> experimental results (on four examples) shows the minor overhead 
> induced by PArrows, which is outweighed by their uniform and general
> interface. 
> 
> Overall Evaluation:
> 
> This paper shows a useful development of a general arrow-based
> interface for specifying parallel computation in Haskell. This
> development is definitely worth doing (and I would wish an earlier
> development of such arrow-based interface).
> 
> The whole paper is clearly written and well structured. It explains
> not only the background and related work to make the paper
> self-contained, but also the motivation and the details of the
> arrow-based language for parallel programming in Haskell and the three
> instance implementations.
> 
> I like the paper and enjoyed reading it, but the *technical*
> contributions of this paper sounds limited with the standard of JFP.
> 
> I agree with the contributions listed in the introduction, but I am
> wondering how challenging it is to have these contributions. Following
> the research work on arrow-interfaces for GUI, inverse computation or
> functional reactive computation, it does not look difficult to define
> such an arrow-based interface for parallel programming for functional
> parallel programming. It is my hope that the authors can clarify the
> challenges and how these challenges are tacked in the later revision.
> 
> Detailed Comments:
> 
> Section 1:
> 
> On attractive point of this arrow-based general interface is that it
> allows to parallel implementation at will. It seems that it would be
> impossible for this switching unless every parallel computing
> primitive is wrapped with the general interface. Is the arrow-based
> language defined in this paper sufficient?  Or in other words, is
> there any parallelism in the current parallel Haskell system (say GhP)
> that cannot be described in this arrow-based interface?
>

We would expect some at least technical difficulties with Accelerate. However, our interface seems to be fairly universal. A very different approach is data parallel programming. Our approach is task parallel.

> Section 3:
> 
> For GpH, can you describe the two primitives par/seq in terms of
> PArrows?

Our approach is directly reversed: we define PArrows in terms of par/seq. However, TODO TODO

> 
> Section 4:
> 
> How general is the new interface language? I see that "parEvalN" in
> Section 4 needs to be extended to deal with "loop" later in Section
> 4. Do we need to extend it again to deal with other cases? Can you
> give a clear definition of the interface language, together with its
> application scope (including discussion what it cannot do)?
> 
> Section 5:
> 
> I got a bit lost when reading the "Future" class, and I cannot fully
> understand why it can support direct communications between
> nodes. What is the relationship between "Future" and "parEvalN"? Is
> "Future" part of PArrows?

It is not part of the class holding parEvalN, but it extends the general interface of PArrows.

> 
> Section 6:
> 
> At the beginning of this section, it says "Now we have developed
> Parallel Arrows far enough to .... While there are many possible
> skeletons to implement, we demonstrate the expressive power of PArrows
> here using four map-based and three toplogical skeletons."  Later in
> Section 6.2, it shows that the original developed PArrows are not good
> enough and should to be extended to deal with "loop". Does this mean
> that PArrows defined in this paper can be the basis for the future
> extension?
> 
> Referee: 3
> 
> Comments to the Author
> # Paper summary
> 
> There are now a large number of Haskell APIs and runtimes for parallel
> computing, from shared memory on a single machine to distributed
> computation. Programmers currently must choose a particular API and
> program to it, making it difficult to experiment with alternative
> parallelism solutions.
> 
> This paper proposes a common API, based on Hughes'
> Arrows. Implementations of this API for GpH, the Par Monad, and Eden
> are given. The authors demonstrate a number of programs written using
> this API, and present some benchmarks to show that using a common API
> does not introduce any significant overhead.
> 
> # Assessment
> 
> The authors present a (mostly) nice looking API for writing parallel
> programs in Haskell, that can be supported by at least three different
> backends, and has low overhead.
> 
> My problem with this paper is that I can't see why Arrows are
> essential. Since the title is "Arrows for Parallel Computation", this
> is a major problem.

> 
> Why don't I think Arrows are essential to this work? The authors
> present the ArrowParallel type class (Sec 4.1, pg 11):
> 
>    class Arrow arr => ArrowParallel arr a b conf where
>       parEvalN :: conf -> [arr a b] -> arr [a] [b]
> 
> and then give instances for this class for GpH, the Par monad, and
> Eden. The former two permit any Arrow instance that is also an
> ArrowChoice, but Eden only permits the (->) and Kleisli
> instances.
> 
> However, as far as I can tell, all the actual example parallel
> programs presented only use the (->) instance (Figure 28 seems to be
> the only serious program presented). Are there any examples of using
> other Arrow instances? Is there any point in using anything other than
> the (->) and Kleisli instances if they don't work with Eden?

It would be possible to creature further special instances for Eden, if those are of interest.

@@

>  make programs go faster, it would seem to me that a particular
>  construct in a paper about parallelism should only appear if it can
>  be shown to actually make programs go faster.

The `pipe` and `ring` skeletons are needed, as they allow for a more explainable and didactic introduction of the `torus`. Basically, `ring` is a warped `pipe`, and `torus` is a `ring` of `ring`s. Hence, the `torus` benchmark is the ultimate test for the whole family of `pipe`, `ring`, and `torus`.

> 
> - A frustrating amount of the code is placed in the appendix (mainly
>  appendix C) and referenced from the main text. It would be much
>  easier to read the paper if all of the code was inlined in the
>  text. In the present state, the reader has to keep flipping to the
>  end to find the implementations of crucial bits of code (e.g.,
>  `lazy`).
> 
> - Throughout, the writing is a bit sloppy. I've highlighted some
>  examples below.
> 
> ## Text comments
> 
> ### Page 2
> 
> Para 1, line 3: "short" => "for short"
> 
> Para 2, line 1: "A key" => "The key"
> 
> Para 3: line 1: "Arrows" => "Arrow"
> 
> Para 3: line 3: "we define a shallow-embedded DSL for Arrows". Aren't
> you defining a shallowly embedded DSL for parallel programming,
> *using* Arrows?
> 
> Para 3: line 6: "very" is vague.

We fixed the above spelling problems, thank you very much.

> 
> Para 4: line 2: is there a missing 'Arrow arr' constraint on the type
> signature?
>

Indeed, we fixed this mistake and apologize.

> Para 6: line 1: this is the first mention of "PArrow" -- it should be
> defined earlier.

We fixed this.

> 
> ### Page 3
> 
> Para 1: line 1: "This has many practical advantages". Are there any
> others beyond being able to switch backend easily?

> 
> Para 6: line 2: "parallelisations" => "execution strategies"

Thank you for this suggestion. We accepted it, the final wording is: "... relatively easy to define alternate execution strategies, including parallel evaluation."

> Para 6: line 7: "is completely orthogonal to our approach". This seems
> a bit defensive. But since you bring it up, then why not explain why
> it is orthogonal?

Accelerate focuses on accessing the GPU and adapts to the notions of Cuda language. We found it hard to combine the concepts of PArrows and Accelerate under one hood. However, we are considering to use PArrows to orchestrate multiple Accelerate instances.

We reason that such an in-depth explanation is going to far in the paper text, hence we omit it.

> 
> Para 8: line 2: Should be a colon after "following"?

Fixed.

> 
> Para 8: line 3: I'm not sure what you mean by "communication-centred"
> in relation to LVars?

> 
> ### Page 4
> 
> The section on Algorithmic skeletons needs a forward reference to
> later in the paper where you discuss how to express skeletons using
> your formalism.
> 
> "Arrows" para: "less restrictive alternative", well, it is less
> restrictive for the writer of the arrow, but not for the user.
>

> "Arrow": line 10: "capsule the computation", do you mean "encapsulate
> the computation"? This whole sentence seems quite vague anyway. I'd
> characterise Liu et al's work as providing a specific implementation
> of arrows that allows for optimisation of arrows with feedback by
> retaining some intensional information about the computation. The
> comparison with you work confuses me: you say that their arrows allow
> parallel composition, but then seem to say that yours is different?
> What is a thing that you can express but they can't?
>

We adjusted the description of Lue et al.'s work.

The difference between our work and Lue et al. is that they would allow for composition of their arrows in parallel. We provide operators to execute a set of arrows in parallel (yielding an arrow again).

WHAT'S THE DIFFERENCE

### 2019-12-27 paper/content/ms.tex
reviewer's comments (2)

Parallel functional languages have a long history of being used for experimenting with novel parallel programming paradigms. Haskell, which we focus on in this paper, has  several mature implementations. We regard here in-depth
Glasgow parallel Haskell or short GpH (its Multicore SMP implementation, in particular), the
\ensuremath{\Conid{Par}} Monad, and Eden, a distributed memory parallel Haskell. These
languages represent orthogonal approaches. Some use a Monad, even if
only for the internal representation. Some introduce additional
language constructs. Section \ref{sec:parallelHaskells} gives a short overview over these languages.

A key novelty in this paper is to use Arrows to represent parallel computations. They seem a natural fit as they can be thought of as a more general function arrow (\ensuremath{\to }) and serve as general interface to computations while not being as restrictive as Monads \citep{HughesArrows}. Section \ref{sec:arrows} gives a short introduction to Arrows.

We provide an Arrows-based type class and implementations for the three above mentioned parallel Haskells.
Instead of 
introducing a new low-level parallel backend to implement our
Arrows-based interface, we define a shallow-embedded DSL for Arrows. This DSL
is defined as a common interface with varying implementations in
the existing parallel Haskells. Thus, we not only define a parallel programming interface in a
novel manner---we tame the zoo of parallel Haskells. We provide a
common, very low-penalty programming interface that allows to switch
the parallel implementations at will. The induced penalty was in the single-digit percent range, with means typically under 2\% overhead in our measurements over the varying cores configuration (Section~\ref{sec:benchmarks}). We have also conducted some initial experiments with HdpH and Frege; those experiments inducate that our parallel Arrows DSL can be relatively easily extended to further languages.

\paragraph{Contributions.}
%
%\olcomment{HIT HERE REALLY STRONG}
%
%\subsection{Impact of parallel Arrows}
%\olcomment{move this to Contributions in the front or something}
We propose an Arrow-based encoding for parallelism based on a new Arrow combinator \ensuremath{\Varid{parEvalN}\mathbin{::}[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]}. A parallel Arrow is still an Arrow, hence the resulting parallel Arrow can still be used in the same way as a potential sequential version. In this paper we evaluate the expressive power of such a formalism in the context of parallel programming.

@@

\paragraph{Parallel Haskells.}
The non-strict semantics of Haskell, and the fact that reduction encapsulates computations as closures, makes it relatively easy to define alternate parallelisations. A range of approaches have been explored, including data parallelism \cite{Chakravarty2007,Keller:2010:RSP:1932681.1863582}, GPU-based approaches \cite{Mainland:2010:NEC:2088456.1863533,obsidian-phd}, software transactional memory \cite{Harris:2005:CMT:1065944.1065952,Perfumo:2008:LST:1366230.1366241}.
The Haskell--GPU bridge Accelerate
\cite{Chakravarty:2011:AHA:1926354.1926358,CMCK14,McDonell:2015:TRC:2887747.2804313}
is completely orthogonal to our approach.
A~good survey of parallel Haskells can be found in~\citet{marlow2013parallel}.

@@

Other task parallel Haskells related to Eden, GpH, and the \ensuremath{\Conid{Par}} Monad include the following.  
HdpH \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} is an extension of \ensuremath{\Conid{Par}} Monad to
heterogeneous clusters. LVish \cite{Kuper:2014:TPE:2666356.2594312} is a communication-centred extension
of \ensuremath{\Conid{Par}} Monad.
We chose to not extend our implementation to LVish and HdpH at the moment. The former does not differ from the original \ensuremath{\Conid{Par}} Monad with regard to how we would have used it in this paper. The latter heavy relies on Template Haskell, which adds complexity to the implementation.

@@

\paragraph{Arrows.}
Arrows were introduced by \citet{HughesArrows} as a less restrictive alternative to Monads, in essence they are a generalised function arrow~\ensuremath{\to }. \citet{Hughes2005} presents a tutorial on Arrows. \citet{jacobs_heunen_hasuo_2009,LINDLEY201197,ATKEY201119} develop theoretical background of Arrows. \citet{Paterson:2001:NNA:507669.507664} introduced a new notation for Arrows. Arrows have applications in information flow research \cite{1648705,LI20101974,Russo:2008:LLI:1411286.1411289}, invertible programming \cite{Alimarine:2005:BAA:1088348.1088357}, and quantum computer simulation \cite{vizzotto_altenkirch_sabry_2006}. But probably most prominent application of Arrows is Arrow-based functional reactive programming, AFRP \cite{Nilsson:2002:FRP:581690.581695,Hudak2003,Czaplicki:2013:AFR:2499370.2462161}.
\citet{Liu:2009:CCA:1631687.1596559} formally define a more special kind of Arrows that capsule the computation more than regular Arrows do and thus enable optimisations. Their approach would allow parallel composition, as their special Arrows would not interfere with each other in concurrent execution. In contrast, we capture a whole parallel computation as a single entity: our main instantiation function \ensuremath{\Varid{parEvalN}} makes a single (parallel) Arrow out of a list of Arrows. \citet{Huang2007} utilise Arrows for parallelism, but strikingly different from our approach. They use Arrows to orchestrate several tasks in robotics. We, however, propose a general interface for parallel programming, while remaining completely in Haskell.
\citet{zhang_session_2019} introduced a session-type framework that generates parallel code. As they stated, their approach to express dataflow with Arrows was inspired by our formalism.

@@

PArrows is an extendable formalism, they can be easily ported to further parallel Haskells while still maintaining interchangeability. It is straightforward to provide further implementations of algorithmic skeletons in PArrows.
	%%\pagebreak

\section{Conclusion}
\label{sec:conclusion}
Arrows are a generic concept that allows for powerful composition
combinators. To our knowledge we are first to represent
\emph{parallel} computation with
Arrows, and hence to show their usefulness for
% Arrows are a useful tool for
composing parallel
programs. We have shown that for a generic and extensible parallel Haskell, we do not have to restrict ourselves to a monadic interface. % Instead, we use Arrows which are a more general concept than Monads, but still allow for powerful composition. Furthermore, we think
We argue that Arrows are better suited to  parallelise
pure functions than Monads,  as the functions are already Arrows and can be used
directly in our DSL.
Arrows are a better fit to parallelise pure code than a monadic solution as regular functions are already Arrows and can be used with our DSL in a more natural way. % We manage to retain this nice property of parallel Haskells such as Eden or GpH (which use pure constructs)
% while still having a generic and composable interface.
We use a non-monadic interface (similar to Eden or GpH) and retain composability.
%
The DSL allows for a direct parallelisation of monadic code via the Kleisli type and additionally allows to parallelise any Arrow type that has an instance for \ensuremath{\Conid{ArrowChoice}}. (Some skeletons require an additional \ensuremath{\Conid{ArrowLoop}} instance.)

>>>>

Parallel functional languages have a long history of being used for experimenting with novel parallel programming paradigms. Haskell, which we focus on in this paper, has  several mature implementations. We regard here in-depth
Glasgow parallel Haskell or for short GpH (its Multicore SMP implementation, in particular), the
\ensuremath{\Conid{Par}} Monad, and Eden, a distributed memory parallel Haskell. These
languages represent orthogonal approaches. Some use a Monad, even if
only for the internal representation. Some introduce additional
language constructs. Section \ref{sec:parallelHaskells} gives a short overview over these languages.

The key novelty in this paper is to use Arrows to represent parallel computations. They seem a natural fit as they can be thought of as a more general function arrow (\ensuremath{\to }) and serve as general interface to computations while not being as restrictive as Monads \citep{HughesArrows}. Section \ref{sec:arrows} gives a short introduction to Arrows.

We provide an Arrow-based type class and implementations for the three above mentioned parallel Haskells.
Instead of 
introducing a new low-level parallel backend to implement our
Arrow-based interface, we define a shallow-embedded DSL using Arrows. This DSL
is defined as a common interface with varying implementations in
the existing parallel Haskells. Thus, we not only define a parallel programming interface in a
novel manner---we tame the zoo of parallel Haskells. We provide a
common, low-penalty programming interface that allows to switch
the parallel implementations at will. The induced penalty was in the single-digit percent range, with means typically under 2\% overhead in our measurements over the varying cores configuration (Section~\ref{sec:benchmarks}). We have also conducted some initial experiments with HdpH and Frege; those experiments inducate that our parallel Arrows DSL can be relatively easily extended to further languages.

\paragraph{Contributions.}
%
%\olcomment{HIT HERE REALLY STRONG}
%
%\subsection{Impact of parallel Arrows}
%\olcomment{move this to Contributions in the front or something}
We propose an Arrow-based encoding for parallelism based on a new Arrow combinator \ensuremath{\Varid{parEvalN}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}}. A parallel Arrow (short: PArrow) is still an Arrow, hence the resulting parallel Arrow can still be used in the same way as a potential sequential version. In this paper we evaluate the expressive power of such a formalism in the context of parallel programming.

@@

\paragraph{Parallel Haskells.}
The non-strict semantics of Haskell, and the fact that reduction encapsulates computations as closures, makes it relatively easy to define alternate execution strategies, including parallel evaluation. A range of approaches have been explored, including data parallelism \cite{Chakravarty2007,Keller:2010:RSP:1932681.1863582}, GPU-based approaches \cite{Mainland:2010:NEC:2088456.1863533,obsidian-phd}, software transactional memory \cite{Harris:2005:CMT:1065944.1065952,Perfumo:2008:LST:1366230.1366241}.
The Haskell--GPU bridge Accelerate
\cite{Chakravarty:2011:AHA:1926354.1926358,CMCK14,McDonell:2015:TRC:2887747.2804313}
is completely orthogonal to our approach.
A~good survey of parallel Haskells can be found in~\citet{marlow2013parallel}.

@@

Other task parallel Haskells related to Eden, GpH, and the \ensuremath{\Conid{Par}} Monad include the following: 
HdpH \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} is an extension of \ensuremath{\Conid{Par}} Monad to
heterogeneous clusters. LVish \cite{Kuper:2014:TPE:2666356.2594312} is a communication-centred extension
of \ensuremath{\Conid{Par}} Monad.
We chose to not extend our implementation to LVish and HdpH at the moment. The former does not differ from the original \ensuremath{\Conid{Par}} Monad with regard to how we would have used it in this paper. The latter heavy relies on Template Haskell, which adds complexity to the implementation.

@@

\paragraph{Arrows.}
Arrows were introduced by \citet{HughesArrows} as a less restrictive alternative to Monads, in essence they are a generalised function arrow~\ensuremath{\to }. \citet{Hughes2005} presents a tutorial on Arrows. \citet{jacobs_heunen_hasuo_2009,LINDLEY201197,ATKEY201119} develop theoretical background of Arrows. \citet{Paterson:2001:NNA:507669.507664} introduced a new notation for Arrows. Arrows have applications in information flow research \cite{1648705,LI20101974,Russo:2008:LLI:1411286.1411289}, invertible programming \cite{Alimarine:2005:BAA:1088348.1088357}, and quantum computer simulation \cite{vizzotto_altenkirch_sabry_2006}. But probably most prominent application of Arrows is Arrow-based functional reactive programming, AFRP \cite{Nilsson:2002:FRP:581690.581695,Hudak2003,Czaplicki:2013:AFR:2499370.2462161}.
\citet{Liu:2009:CCA:1631687.1596559} formally define a more special
% kind of Arrows that encapsulate the computation more than regular
% Arrows do and thus enable optimisations.
implementation of Arrows that allows for better optimisations.
Their approach would allow parallel composition, as their special Arrows would not interfere with each other in concurrent execution. In contrast, we capture a whole parallel computation as a single entity: our main instantiation function \ensuremath{\Varid{parEvalN}} makes a single (parallel) Arrow out of a list of Arrows. \citet{Huang2007} utilise Arrows for parallelism, but strikingly different from our approach. They use Arrows to orchestrate several tasks in robotics. We, however, propose a general interface for parallel programming, while remaining completely in Haskell.
\citet{zhang_session_2019} introduced a session-type framework that generates parallel code. As they stated, their approach to express dataflow with Arrows was inspired by our formalism.

@@

PArrows is an extendable formalism that can be easily ported to further parallel Haskells while still maintaining interchangeability. It is straightforward to provide further implementations of algorithmic skeletons in PArrows.
	%%\pagebreak

\section{Conclusion}
\label{sec:conclusion}
Arrows are a generic concept that allows for powerful composition
combinators. To our knowledge we are first to represent
\emph{parallel} computation with
Arrows, and hence to show their usefulness for
% Arrows are a useful tool for
composing parallel
programs. We have shown that for a generic and extensible parallel Haskell, we do not have to restrict ourselves to a monadic interface. % Instead, we use Arrows which are a more general concept than Monads, but still allow for powerful composition. Furthermore, we think
% We argue that Arrows are better suited to  parallelise
% pure functions than Monads,  as the functions are already Arrows and can be used
% directly in our DSL.
Arrows are a better fit to parallelise pure code than a monadic solution as regular functions are already Arrows and can be used with our DSL in a more natural way. % We manage to retain this nice property of parallel Haskells such as Eden or GpH (which use pure constructs)
% while still having a generic and composable interface.
We use a non-monadic interface (similar to Eden or GpH) and retain composability.
%
The DSL allows for a direct parallelisation of monadic code via the Kleisli type and additionally allows to parallelise any Arrow type that has an instance for \ensuremath{\Conid{ArrowChoice}}. (Some skeletons require an additional \ensuremath{\Conid{ArrowLoop}} instance.)
